Neural Networks Loss Landscape Convergence in Different Low-Dimensional Spaces

### 2025-03-20 paper/main.tex
Universal commits presents...

\begin{definition}
  The empirical loss function for the first $k$ elements is:
  $$\mathcal{L}_k(\bw) = \frac1k \sum\limits_1^k \ell_i(\bw), \,\, \mathcal{L}(\bw) := \mathcal{L}_m(\bw).$$
\end{definition}

@@

\begin{definition}
  To calculate the overall loss landscape changing, one has to integrate the absolute difference for the entire parameter space.
  We define \textbf{weighted difference} as:
  $$\Delta_{k} = \int \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right)^2 p(\bw) dw,$$
  where $p(\bw)$ describes the priority of the particular parameter points so we can make $p(\bw)$ have higher values
  next to the local minima.
\end{definition}

@@

\section{Experiments}\label{sec:exp}

To verify the theoretical estimates obtained, we conducted a detailed empirical study...

>>>>

\begin{definition}
  The empirical loss function for the first $k$ elements is:
  $$\mathcal{L}_k(\bw) = \frac1M \sum\limits_1^k \ell_i(\bw), \,\, \mathcal{L}(\bw) := \mathcal{L}_m(\bw).$$
\end{definition}

@@

\begin{definition}
  To calculate the overall loss landscape changing, one has to integrate the absolute difference for the entire parameter space.
  We define $\mathbf{\Delta}$-\textbf{function} (delta-function) as:
  $$\Delta_{k} = \int \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right)^2 p(\bw) dw,$$
  where $p(\bw)$ describes the priority of the particular parameter points so we can make $p(\bw)$ have higher values
  next to the local minima.
\end{definition}

@@

Let $\mathbf{V} = \{\mathbf{e}_1, \dots, \mathbf{e}_d\}$ be eigenvectors of $\mathbf{H}$. We select $d$ of them and project parameters space
into a subspace on these $d$ vectors. These vectors should represent maximum eigenvalues of $\mathbf{H}$ as these directions
have the most impact to the loss function difference.
\section{Experiments}\label{sec:exp}

To verify the theoretical estimates obtained, we conducted a detailed empirical study.
In this section, we present experiments aimed at analyzing how the loss landscape evolves as the size of the training
set increases.

\paragraph{Data Preparation.}
We use the MNIST dataset, which consists of $60{,}000$ training images and $10{,}000$ test
images of handwritten digits (0--9). Each image is $28\times28$ pixels, grayscale.

\paragraph{Model Architecture.}
We use the following neural network models and preparations:
\begin{itemize}
  \item A multilayer perceptron (MLP) with different number of hidden layers (ReLU activations).
  \item The choice of architecture remains consistent across all experiments.
\end{itemize}

The first two are preliminary experiments, conducted as illustrative examples on the MNIST dataset,
while the others focusing on Hessian-based projections and show connections between eigenvalues and loss function convergence.
That is our primary aim --- to check our theoretical estimates in-deal.

\subsection{Visualizing the Loss Landscape in a Random Two-Dimensional Subspace}
The idea of this initial experiment is to get an intuitive sense of how the loss function’s landscape shifts with
different training-set sizes. To do this, we pick two random directions $\mathbf{v_1}$ and $\mathbf{v_2}$ in the parameter
space and project the network’s parameters onto the subspace they span.

\paragraph{Procedure}
\begin{enumerate}
  \item \textbf{Training on different sizes.} Train a neural network on increasingly larger subsets.
  \item \textbf{Parameter snapshot.} After training, fix the final (or near-final) weights $\mathbf{w}^*$.
  \item \textbf{Subspace grid.} Form a grid of points $\{\mathbf{w}^* + \alpha \mathbf{v}_1 + \beta \mathbf{v}_2\}$ around $\mathbf{w}^*$, where $\alpha$ and $\beta$ vary in a chosen range.
  \item \textbf{Loss evaluation.} Compute and plot the loss at each grid point, producing a 3D visualization.
\end{enumerate}

For smaller $N$, the landscape exhibits sharper fluctuations. As $N$ grows, the landscape often appears smoother and more stable.

\begin{figure}[!htbp]
  \hspace*{-11.5cm}
  \subfloat{\includegraphics[width=2\textwidth]{img/Ls_16.pdf}}
  \caption{Loss function landscape}
  \label{fig:base_loss}
\end{figure}

\begin{figure}[!htbp]
  \hspace*{-11.5cm}
  \subfloat{\includegraphics[width=2\textwidth]{img/Ld_10.pdf}}
  \caption{Loss function difference $(\mathcal{L}_{k} - \mathcal{L}_{k-1})^2$ landscape}
  \label{fig:base_loss_diff}
\end{figure}

\subsection{Visualizing delta-function}

Now we dig deeper and visualize our $\Delta_k$ function to clearly understand the difference in the loss function when
moving from a model trained on $k-1$ samples to one trained on $k$ samples. To do this we sample a bunch of random gaussian
vectors (all components are normally distributed) and calculate a
math expectation of modulus of the loss function difference.
Directly computing this expectation over the entire distribution of random directions $\mathbf{d}$
is impossible, so we employ a Monte Carlo approximation: using Monte-Carlo method (we take the mean value of loss function
on sampled gaussian shifted points):

$$\Delta_k = \mathbb{E}_{p(\bw)} \left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right)^2 \approx
  \frac{1}{k} \sum_{t=1}^{k} \left( \mathcal{L}_k(\mathbf{w}^* + \mathbf{d}_t) - \mathcal{L}_{k-1}(\mathbf{w}^* + \mathbf{d}_t) \right)^2$$

where $\mathbf{d}_t \sim \mathcal{N}(0,I_d)$ are $K$ random gaussian samples drawn from a multivariate normal distribution
with zero mean and identity covariance from a subspace of rank $d$. Specifically, we proceed as follows:

\begin{enumerate}
  \item \textbf{Sampling Directions.} Generate $K$ independent directions $\mathbf{d}_1, \dots, \mathbf{d}_M$
        from $\mathcal{N}(0,I_d)$.
  \item \textbf{Loss Evaluation.} For each direction $\mathbf{d}_t$, evaluate the loss at
        $\mathbf{w}^* + \mathbf{d}_t$ and
        $$\left( \mathcal{L}_{k}(\bw) - \mathcal{L}_{k-1}(\bw) \right)^2 $$
  \item \textbf{Averaging.} Average these values values over $k = 1, \dots, K$ to approximate $\Delta_k$.
\end{enumerate}

By increasing $K$, we obtain a more reliable estimate of the expected difference in loss landscapes.
This Monte Carlo approach is straightforward to implement and does not require explicit knowledge of the underlying loss
function's distribution.

The $\Delta$-function generally decreases as $N$ increases, suggesting that the landscape differences between neighbouring
sizes become smaller.
Beyond a certain point, $\Delta$-values are nearly negligible, indicating that adding more data has minimal impact on
the local landscape in these directions.

\noindent
\textbf{Conclusion from Preliminary Experiments.}
These initial experiments demonstrate that the loss landscape becomes progressively more stable with increasing $N$.
However, the directions used here are either random. This motivates a more rigorous approach in the \emph{main experiment}
below, where we focus specifically on the most critical directions of curvature --- those corresponding to the top
eigenvalues of the Hessian.

\subsection{Main Experiment: Projection onto Dominant Hessian Eigenvectors}

\paragraph{Rationale}
From prior work \cite{sagun2018empirical}, we know that only a few Hessian eigenvectors (those associated with the largest
eigenvalues) tend to capture the most significant curvature directions. By projecting onto these principal directions,
we can more precisely measure how the landscape the loss surface changes as $k$ increases.

\paragraph{Procedure}
\begin{enumerate}
  \item \textbf{Hessian approximation.} For each trained model, approximate
        the Hessian $\mathbf{H}$ or its dominant eigenvalues/eigenvectors. Techniques ???
  \item \textbf{Eigen-decomposition.} Identify the top $d$ eigenvectors $\{\mathbf{e}_1, \dots, \mathbf{e}_d\}$ with the
        largest eigenvalues.
  \item \textbf{Subspace exploration.} For each model, calculate a $\Delta$ fuction via the same way as in the second
        experiment (using Monte-Carlo) and appy visualization/.
\end{enumerate}

\paragraph{Goal and Expected Outcome}
\begin{itemize}
  \item \textbf{Refined understanding.} By focusing on the directions of largest curvature, we expect a clearer
        measure of how the landscape of the loss surface changes with additional data.
  \item \textbf{Identifying the threshold.} This approach should yield a more accurate estimate of the point at which the
        landscape effectively stabilizes, thus helping to pinpoint a minimum viable dataset size.
  \item \textbf{Connect with theoretical results.} By applying experiments we check our calculations above on the reality.
\end{itemize}

\begin{figure}[!htbp]
  \hspace*{-0.7cm}
  \subfloat{\includegraphics[width=1.15\textwidth]{img/D_32.pdf}}\\
  \caption{$\Delta$-function visualization. `Sigma` parameter refers to the variance of sampled gaussian vectors, `dim' 
  refers to the rank of the subspace.}
  \label{fig:base_delta}
\end{figure}
